Type Evolution in Programming Systems

### 2024-05-28 paper.tex
Work in progress on conf organizer

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{tikz-cd}
\usepackage{float}

@@

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}\setlength{\parindent}{0em}{\firamedium Framework perspective.}\quad #2
\end{tcolorbox}}

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.6\textwidth]{figures/arrows.png}
\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
  in data and code whose shape is defined by type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

\frameworkbox{
The evolving \emph{type} in this case is the \texttt{State} type, \emph{code} is the
program logic implemented in \texttt{update} and \texttt{render} and \emph{data} is the current
state. To support live programming, a local type change requires synchronizing local data to
match the new type, and also a change in code so that it continues working against the new type
structure. In a non-local scenario, programmers would likely want to synchronize types and code,
but not necessarily data. A collaborative programming system would need to support decentralized
and possibly divergent type evolution, but without data synchronization.
}

@@

\frameworkbox{
In a programming system, the \emph{types} to which entity evolution applies can be data tables,
but also objects or JSON. When the type evolves, the system needs to update corresponding
live \emph{data} and \emph{code}. In a non-local scenario, entity evolution
typically concerns permanent data. As a result, the system needs to synchronize types,
data and code across all replicas. Doing so in a way that allows divergence is discussed
in the next section.
}

@@

@@

\begin{figure}
\begin{adjustwidth}{-5.5em}{-5.5em}
\centering
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Adele Goldberg, adele@xerox.com
    \item Margaret Hamilton, hamilton@mit.com
    \item Betty Jean Jennings, betty@rand.com
   \end{itemize}
   \vspace{1em}
   }}
   \caption{Initial state of the document.}
   \label{fig:conf-orig}
\end{subfigure}
\hfill
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Ada Lovelace, ada@rsoc.ac.uk
    \item Adele Goldberg, adele@xerox.com
    \item Betty Jean Jennings, betty@rand.com
    \item Margaret Hamilton, hamilton@mit.com
   \end{itemize}
   }}
   \caption{Added a speaker and sorted the list.}
   \label{fig:conf-add}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \begin{tabular}{| l | l | l |}
     \hline
     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
     \hline \hline
     Adele Goldberg & adele@xerox.com & TP\\
     \hline
     Margaret Hamilton & hamilton@mit.com & JE\\
     \hline
     Betty Jean Jennings & betty@rand.com & JE\\
     \hline
   \end{tabular}
   \\~
   }}
   \caption{Refactored list into a table.}
   \label{fig:conf-tab}
\end{subfigure}
\hfill
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \begin{tabular}{| l | l | l |}
     \hline
     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
     \hline \hline
     Ada Lovelace & ada@rsoc.ac.uk & \\
     \hline
     Adele Goldberg & adele@xerox.com & TP\\
     \hline
     Betty Jean Jennings & betty@rand.com & JE\\
     \hline
     Margaret Hamilton & hamilton@mit.com & JE\\
     \hline
   \end{tabular}
   }}
   \caption{Resulting document after change merging.}
   \label{fig:conf-fin}
\end{subfigure}
\end{adjustwidth}
\vspace{0.5em}
\caption{Conference organizer. Initial version of the document with two independent changes
and the final version of the document after the two changes are merged.}
\label{fig:conf}
\vspace{0.5em}
\end{figure}

\section{Conference Organizer: Merging edits in computational documents}
In this section, we consider challenges related to working with a document-based format akin to HTML.
We discuss two variants of the format. In the first format, a document
contains only data. In the second variant, the format is extended with computed values.
A computed value is a node in the document whose value is specified by a formula akin to those
used in spreadsheets. The equations can refer to other nodes in the document, possibly using
an absolute or a relative path and may involve a range of built-in functions, for example to
count the number of elements or to sum a sequence of numerical values.

We assume that the document can be
concurrently edited by multiple users in the same fashion as in local-first software, i.e.,
users may edit a local copy that then needs to be synchronized with changes made by other users.
The edits to the document can be done in a rich-text editor, or by using a more structure-aware
structure editor that can, for example, record changes as a sequence of high-level edit operations.

The concrete challenges are centered around the problem of collaborative conference planning
as illustrated in Figure~\ref{fig:conf}. A number of co-organizers are planning a conference
that will feature talks by several invited speakers. They need to agree on the speaker list,
contact speakers and calculate the conference budget.

\frameworkbox{
In this scenario, the only transient \emph{state} is the result of computed values. The content
of the document corresponds to the \emph{data} pace layer. A change in the data layer may invalidate
the computed value at the state layer. Formulas correspond to the \emph{code} layer and changing them,
again, invalidates the state. The \emph{schema} pace layer is implicit. A change to the document that
modifies content is a mere data change, but restructuring the document is an implicit schema
change. Such schema change that may affect the code pace layer.

\paragraph{Applications.}
One of the earliest systems to combine documents with computational elements
is Boxer \cite{diSessa86}. The challenges in this section apply to a wide range of
systems that can be seen as its successors. The specific scenario discussed here is
inspired by the use of Notion~\cite{notion}, which is a collaborative structured document
editing system. Treating documents as end-user programs has recently
been explored in Potluck~\cite{Litt2023}, which is local-only but includes formulas similar
to those in our challenge.

The challenges in this section are also relevant to work on computational scientific notebooks.
Jupyter \cite{Kluyver2016} uses a simple representation as a sequence of cells and does not have
a mechanism for embedding data in the document and referring to it, but numerous
project such as Nextjournal \cite{Nextjournal21} advance the format. The idea of producing
interactive, computational secientific documents is an active research area. Recent contributions
include Nota~\cite{Crichton2021} and Living Papers \cite{Heer2023}.

\challenge{Merging structured document edits}

First, the organizers need to agree on a list of speakers to invite, coordinate who contacts
whom and track the acceptance of invitations. They start with a document shown in
Figure~\ref{fig:conf-orig}. The challenge is to merge document edits that are done locally
and independently by two co-organizers. Specifically, consider the following two edits:

\begin{enumerate}
\item The first organizer adds an additional speaker to the list and sorts the list of speakers
  alphabetically by their first name. (Figure~\ref{fig:conf-add})
\item The second organizer refactors the list into a table. They split the single textual value
  into a name and an email (using a comma as the separator) and add an additional column for
  tracking which of the organizers should contact the speaker.
\end{enumerate}

\noindent
The system should be able to merge the changes and produce a final table shown in
Figure~\ref{fig:conf-fin}. The resulting table needs to include the additional speaker
created by the first organizer, use the order specified by the first organizer and use
the format defined by the second organizer. The newly added speaker should be reformatted
into the new format. For the newly added ``Organizer'' column, the second user may
specify default value or the newly added row may use an empty value.

\frameworkbox{
In this challenge, the change done by the first organizer is at the \emph{data} layer,
while the change done by the second organizer spans the \emph{schema} and \emph{data} layers.
In implementing the schema change, the second organizer transforms the affected document data.
The challenge is to apply the same transformation to the data independently added by the
second organizer.
}

\begin{figure}
\begin{adjustwidth}{-5.5em}{-5.5em}
\centering
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Adele Goldberg, adele@xerox.com
    \item Margaret Hamilton, hamilton@mit.com
    \item Betty Jean Jennings, betty@rand.com
   \end{itemize}
   \vspace{1.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}
   \caption{Initial state of the document with formulas.}
   \label{fig:conf-budget}
\end{subfigure}
\hfill
\begin{subfigure}[b]{23em}\vspace{0pt}
   \fbox{\parbox{22.6em}{
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Resulting document after schema change merging.}
   \label{fig:conf-finfin}
\end{subfigure}
\caption{Conference organizer. Formulas to calculate the budget have been added to the
original document (Figure~\ref{fig:conf-orig}) and those have to be merged with the
refactoring of a list into a table (Fig~\ref{fig:conf-tab}). Formulas in the final version
are updated to match with the new document structure.}
\label{fig:conf-calc}
\end{adjustwidth}
\end{figure}

\frameworkbox{
In this challenge, the change done by the first organizer is at the \emph{code} layer, while
the change done by the second organizer is both at the \emph{data} layer and, implicitly,
at the \emph{schema} layer. A change solely at the \emph{data} layer would not affect code,
but a change at the \emph{schema} layer requires adapting some aspects of the code.
}

\frameworkbox{
In the \emph{divergence} model, it may be expected that users will want to adopt all changes
done at the \emph{data} pace layer. They may selectively choose some of the changes at the
\emph{code} and \emph{schema} layers. The challenge is maintaining the same data across multiple
program variants and merging changes done to the data based on different document schema.
}

\frameworkbox{
In the live extension, we also consider the transient \emph{state} pace layer. A system should
automatically detect when a change at the \emph{data} or \emph{code} layers affects the computed
value and forces it to be recomputed. Interestingly, a change at the \emph{schema} layer may not
always affect the \emph{state} layer, even if it schema change requires modifying formulas that
compute the state.
}

@@

\section{Related Work}

>>>>

\usepackage{csquotes}
\usepackage{changepage}
\usepackage{multicol}
\usepackage{ccicons}
\usepackage[many]{tcolorbox}
\usepackage{tikz-cd}
\usepackage{float}
\usepackage{wrapfig}

@@

\DeclareRobustCommand{\frameworkbox}[2][gray!15]{
\begin{tcolorbox}[breakable,left=3pt,right=3pt,top=3pt,bottom=3pt,colback=#1,colframe=#1,parbox=false,
  width=\dimexpr\textwidth\relax,enlarge left by=0mm,boxsep=5pt,arc=0pt,enlarge top by=0.5em,%enlarge bottom by=0.0em,
  outer arc=0pt]\setlength{\parskip}{0.5em}#2
\end{tcolorbox}}
\newcommand{\frameworkboxtitle}[1]{\setlength{\parindent}{0em}{\firamedium #1.}\quad}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.175]{figures/diag-data-code.png}
\caption{When a type changes from $T_1$ to $T_2$, a corresponding change needs to take place
  in data and code whose shape is defined by type. The challenge is finding a suitable
  corresponding transformation (semi-)automatically.}
\label{fig:change}
\end{figure}

@@

\frameworkbox{
\begin{wrapfigure}{r}{14em}
\vspace{1em}
\includegraphics[scale=0.175]{figures/diag-both.png}
\end{wrapfigure}
\frameworkboxtitle{Code-data Co-evolution}
In this case study, the evolving \emph{type} is the \texttt{State} type. When the type
evolves, corresponding change needs to be applied to both \emph{code} and \emph{data}.
The former is the program logic implemented in \texttt{update} and \texttt{render},
while the latter is the current state (a live value of type \texttt{State}). The scenario
illustrates that code and data need to co-evolve as illustrated by a revised diagram on the right.
}

@@

\frameworkbox{
\begin{wrapfigure}{r}{8em}
\includegraphics[scale=0.175]{figures/diag-variants.png}
\end{wrapfigure}
\frameworkboxtitle{Primitive Type Transformations}
In this example, \emph{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

@@

% ==================================================================================================
% DOCUMENTS
% ==================================================================================================

\section{Evolution of Computational Documents}
A different kind of type evolution is present in programming systems based on documents.
The idea, dating back to Boxer \cite{diSessa86}, is to represent programs as documents
with embedded computed values. In our model, a computed value is a document node whose value is
specified by a simple formula (akin to those in Potluck~\cite{Litt2023} or spreadsheets) that
may refer to other nodes of the document. We assume that documents can be concurrently edited
by multiple users in a local-first manner, i.e., local edits are later synchronized with peers.

Our concrete example is collaborative conference planning, illustred in Figure~\ref{fig:conf}
and inspired by a typical use of Notion~\cite{notion}. A number of co-organizers are planning
a conference that will feature talks by several invited speakers. They need to agree on the
speaker list, contact speakers and calculate the conference budget.
%
The ideas in this section also apply to work on computational scientific notebooks
that combine code and data such as Jupyter \cite{Kluyver2016}. We hope to contribute to ongoinig
work on advancing the format as done by Nextjournal \cite{Nextjournal21}, Nota~\cite{Crichton2021}
and Living Papers~\cite{Heer2023}.

\begin{figure}
\begin{adjustwidth}{-2.5em}{-2.5em}
\centering
\begin{subfigure}[b]{18em}\vspace{0pt}
   \fbox{\parbox{17.6em}{\small
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \begin{itemize}
    \item Ada Lovelace, ada@rsoc.ac.uk
    \item Adele Goldberg, adele@xerox.com
    \item Betty Jean Jennings, betty@rand.com
    \item Margaret Hamilton, hamilton@mit.com
   \end{itemize}
   }}
   \caption{Jonathan adds a new speaker and sorts the list.}
   \label{fig:conf-add}
\end{subfigure}
\hfill
\begin{subfigure}[b]{22em}\vspace{0pt}
   \fbox{\parbox{21.6em}{\small
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}
   \vspace{0.3em}

   \begin{tabular}{| l | l | l |}
     \hline
     \textbf{Name} & \textbf{Email} & \textbf{Who} \\
     \hline \hline
     Adele Goldberg & adele@xerox.com & TP\\
     \hline
     Margaret Hamilton & hamilton@mit.com & JE\\
     \hline
     Betty Jean Jennings & betty@rand.com & JE\\
     \hline
   \end{tabular}
   \\~
   }}
   \caption{Tomas refactors the list into a table.}
   \label{fig:conf-tab}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{18em}\vspace{0pt}
   \fbox{\parbox{17.6em}{\small
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   ~\\[-0.15em]
   \textbf{Invited speakers}
   \begin{itemize}
    \item Adele Goldberg, adele@xerox.com
    \item Margaret Hamilton, hamilton@mit.com
    \item Betty Jean Jennings, betty@rand.com
   \end{itemize}
   \vspace{1.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/ul[id='speakers']/li)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}
   \caption{Tijs adds formulas to compute the total cost.}
   \label{fig:conf-budget}
\end{subfigure}
\hfill
\begin{subfigure}[b]{22em}\vspace{0pt}
   \fbox{\parbox{21.6em}{\small
   \textbf{PROGRAMMING CONFERENCE 2023}\\
   \textbf{Invited speakers}

   \vspace{0.5em}
   \textbf{Conference budget}\\
   Travel cost per speaker:\\
   \hspace*{2em} \$1200 \\
   Number of speakers: \\
   \hspace*{2em} \texttt{=COUNT(/table[id='speakers']/tbody/tr)} \\
   Travel expenses:\\
   \hspace*{2em} \texttt{=/dl/dd[0] * /dl/dd[1]}
   }}\caption{Geoffrey adopts changes from all three co-organizers.}
   \label{fig:conf-finfin}
\end{subfigure}
\end{adjustwidth}
\caption{
Initial version of the document (not shown) contains a list of three speakers (Goldberg, Hamilton,
Jennings). The first challenge involves merging two independent edits to the data
(Fig.~\ref{fig:conf-add} and Fig.~\ref{fig:conf-tab}); the second merging structural change
with code (Fig.~\ref{fig:conf-budget} and Fig.~\ref{fig:conf-tab}). Formulas in the final version
(Fig.~\ref{fig:conf-finfin}) are updated to match the new document structure.}
\label{fig:conf}
\end{figure}

\challenge{Merging Structured Document Edits}

The conference organizers first need to agree on a list of speakers to invite, coordinate who
contacts whom and track the acceptance of invitations. They start with a list of three
speakers (not shown). The first challenge is to merge document edits done locally
by Jonathan and Tomas and change the list and the structure of the document:

\begin{enumerate}
\item Jonathan adds an additional speaker to the list and sorts the list of speakers
  alphabetically by their first name. (Figure~\ref{fig:conf-add})
\item Tomas refactors the list into a table. He splits the single textual value
  into a name and an email (using a comma as the separator) and adds an additional column for
  tracking which of the organizers should contact the speaker.
\end{enumerate}

\noindent
The system should be able to merge the changes and produce a table containing the new data
in a tabular format (not shown). The result needs to include the additional speaker
created by Jonathan, use the order specified by Jonathan, but use the format defined by
Tomas. The newly added speaker should be reformatted into the new format. For the newly added
``Organizer'' column, Tomas may specify default value or the newly added row may use an
empty value.

\frameworkbox{
\frameworkboxtitle{Framework perspective}
In this scenario, the only transient \emph{state} is the result of computed values. The content
of the document corresponds to the \emph{data} pace layer. A change in the data layer may invalidate
the computed value at the state layer. Formulas correspond to the \emph{code} layer and changing them,
again, invalidates the state. The \emph{schema} pace layer is implicit. A change to the document that
modifies content is a mere data change, but restructuring the document is an implicit schema
change. Such schema change that may affect the code pace layer.

\frameworkbox{
\frameworkboxtitle{Framework perspective}
In this challenge, the change done by the first organizer is at the \emph{data} layer,
while the change done by the second organizer spans the \emph{schema} and \emph{data} layers.
In implementing the schema change, the second organizer transforms the affected document data.
The challenge is to apply the same transformation to the data independently added by the
second organizer.
}

\frameworkbox{
\frameworkboxtitle{Framework perspective}
In this challenge, the change done by the first organizer is at the \emph{code} layer, while
the change done by the second organizer is both at the \emph{data} layer and, implicitly,
at the \emph{schema} layer. A change solely at the \emph{data} layer would not affect code,
but a change at the \emph{schema} layer requires adapting some aspects of the code.
}

\frameworkbox{
\frameworkboxtitle{Framework perspective}
In the \emph{divergence} model, it may be expected that users will want to adopt all changes
done at the \emph{data} pace layer. They may selectively choose some of the changes at the
\emph{code} and \emph{schema} layers. The challenge is maintaining the same data across multiple
program variants and merging changes done to the data based on different document schema.
}

\frameworkbox{
\frameworkboxtitle{Framework perspective}
In the live extension, we also consider the transient \emph{state} pace layer. A system should
automatically detect when a change at the \emph{data} or \emph{code} layers affects the computed
value and forces it to be recomputed. Interestingly, a change at the \emph{schema} layer may not
always affect the \emph{state} layer, even if it schema change requires modifying formulas that
compute the state.
}

@@

@@
